Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks

### 2025-06-02 inst/paper/main.tex
update intro and methods from discussion with Adam and update .bib, add model schematic figure and remove tikz figure

\title{Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks}
\author{Joshua W. Lambert}
\date{}

@@

The likelihood of an outbreak being controlled, in other words being locally eliminated, is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case. If the reproduction number is greater than 1, the incidence of infections is likely to increase, and intervention is requireed to cease transmission in a population. If a pathogen is subcritical (i.e. $R < 1$), incidence will decline and eventually stop spreading, but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}. Individual-level heterogeneity in transmission also influences the possibility of a pathogen going extinct from stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}. Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being controlled (however a small number of outbreaks may become large due to superspreading). In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}. \\

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as isolation and contact tracing, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with short generation times, swift and effective interventions are required to achieve containment. Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}. \\

Isolation of infected individuals aims to break the chain of transmission by preventing further infection of susceptible individuals. The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal. This is challenging for most pathogens because infectiousness preceeds prodromal symptoms. Contact tracing is the process of finding the contacts of infected individuals. Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic. Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed. Isolation and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}. Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\

It has become widely accepted that influenza viruses cannot be controlled by isolating individuals at symptom onset and contact tracing, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have low levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

@@

The time between an infector being infected and then infecting an infectee (i.e. generation time) is indirectly computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, infectiousness onset and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}. This formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is accurately estimated but the generation time (or serial interval) is estimated with more uncertainty or unknown. This controls the proportion of the infections that occur prior to, or after, the symptom onset of an infector. The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at two, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution). In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector. \\

@@

\begin{figure}[ht]
\centering
{\sffamily
  \begin{tikzpicture}[every node/.style={font=\tiny}]
  % Labels above the A row
  \node at (0,0.5) {Infected};
  \node at (2,0.5) {Symptomatic};
  \node at (5,0.5) {Isolated};

  % Top row: A circles with red background
  \fill[red!30] (0,0) circle (5pt);
  \draw (0,0) circle (5pt);
  \node at (0,0) {A};

  \fill[red!30] (2,0) circle (5pt);
  \draw (2,0) circle (5pt);
  \node at (2,0) {A};

  \fill[red!30] (5,0) circle (5pt);
  \draw (5,0) circle (5pt);
  \node at (5,0) {A};

  % Horizontal line between A circles
  \draw (0,0) -- (2,0);
  \draw (2,0) -- (5,0);

  % Individual A delays
  \draw[<->] (0.5,-0.3) -- (1.5,-0.3);
  \node at (1,-0.5) {Incubation period};
  \draw[<->] (2,-0.3) -- (5,-0.3);
  \node at (3.5,-0.5) {Onset-to-isolation};

  % Labels above the B row
  \node at (4,-1.5) {Infected};
  \node at (6,-1.5) {Symptomatic};
  \node at (8,-1.5) {Isolated};

  % Middle row: B circles with blue background
  \fill[blue!30] (4,-2) circle (5pt);
  \draw (4,-2) circle (5pt);
  \node at (4,-2) {B};

  \fill[blue!30] (6,-2) circle (5pt);
  \draw (6,-2) circle (5pt);
  \node at (6,-2) {B};

  \fill[blue!30] (8,-2) circle (5pt);
  \draw (8,-2) circle (5pt);
  \node at (8,-2) {B};

  % Horizontal line between B circles
  \draw (4,-2) -- (6,-2);
  \draw (6,-2) -- (8,-2);

  % Labels above the C row
  \node at (4.5,-3.5) {Infected};
  \node at (7.5,-3.5) {Symptomatic};
  \node at (9,-3.5) {Isolated};

  % Bottom row: C circles with green background
  \fill[green!30] (4.5,-4) circle (5pt);
  \draw (4.5,-4) circle (5pt);
  \node at (4.5,-4) {C};

  \fill[green!30] (7.5,-4) circle (5pt);
  \draw (7.5,-4) circle (5pt);
  \node at (7.5,-4) {C};

  \fill[green!30] (9,-4) circle (5pt);
  \draw (9,-4) circle (5pt);
  \node at (9,-4) {C};

  % Horizontal line between C circles
  \draw (4.5,-4) -- (7.5,-4);
  \draw (7.5,-4) -- (9,-4);

  % Labels above the D row
  \node at (8,-5.5) {Infected};

  % Individual C circles with green background
  \fill[teal!30] (8,-6) circle (5pt);
  \draw (8,-6) circle (5pt);
  \node at (8,-6) {D};

  % arrow for asymptomatic D individual
  \draw[->] (8,-6) -- (11,-6); % arrow for asymptomatic

  % Labels above the C row
  \node at (8.75,-7.5) {Infected};
  \node at (10,-7.5) {Symptomatic};
  \node at (11,-7.5) {Isolated};

  % E individual
  \fill[violet!30] (8.75,-8) circle (5pt);
  \draw (8.75,-8) circle (5pt);
  \node at (8.75,-8) {E};

  \fill[violet!30] (10,-8) circle (5pt);
  \draw (10,-8) circle (5pt);
  \node at (10,-8) {E};

  \fill[violet!30] (11,-8) circle (5pt);
  \draw (11,-8) circle (5pt);
  \node at (11,-8) {E};

  % Horizontal line between E circles
  \draw (8.75,-8) -- (10,-8);
  \draw (10,-8) -- (11,-8);

  % Vertical dashed lines
  \draw[dashed] (4,0) -- (4,-2);   % A to B
  \draw[dashed] (4.5,-2) -- (4.5,-4);  % B to C
  \draw[dashed] (8,-4) -- (8,-6);  % C to D
  \draw[dashed] (8.75,-4) -- (8.75,-8);  % C to E

  \end{tikzpicture}
}
\caption{A schematic of the branching process model with isolation.}
\label{fig:ringbp-model}
\end{figure}

>>>>

\title{Evaluating isolation and contact tracing to control pandemic potential influenza outbreaks}
\author{Joshua W. Lambert, Adam J. Kucharski}
\date{}

@@

The likelihood of an outbreak causing sustained transmission in a population is in part based on the epidemiological characteristics of the pathogen, and in part on the inventions implemented, both pharmaceutical and non-pharmaceutical. A key epidemiological metric to quantify the spread of an infectious disease is the reproduction number ($R$), defined as the average number of secondary cases  per primary case. If the reproduction number is greater than 1, the incidence of infections is likely to increase, and intervention is requireed to cease transmission in a population. If a pathogen is subcritical (i.e. $R < 1$), incidence will decline and eventually stop spreading, but can still cause transient outbreaks \citep{farringtonDistributionTimeExtinction1999}. Individual-level heterogeneity in transmission also influences the possibility a pathogen will fail to maintain transmission due to stochasticity in the transmission process \citep{lloyd-smithSuperspreadingEffectIndividual2005}. Higher individual-level variation for a given reproduction number leads to, on average, more outbreaks being unsustained (however a small number of outbreaks may become large due to superspreading). In addition, the likelihood of containment is reduced --- assuming $R > 1$ --- if there are multiple introductions into a populations \citep{whittleOutcomeStochasticEpidemicA1955, kucharskiEarlyDynamicsTransmission2020}. \\

A range of interventions can be deployed to reduce or prevent the transmission of infectious diseases, broadly classified into pharmaceutical and non-pharmaceutical interventions (NPIs). Early in an outbreak --- especially in the case of a novel pathogen --- NPIs are often the first line of defence, as pharmaceutical options may not yet be available. These interventions vary in scope and specificity, from targeted measures such as isolation and contact tracing, to broader actions like school closures, bans on gatherings, and regional or national lockdowns (Huag et al., 2020). The success of such interventions in containing an outbreak before it becomes sustained and widespread depends on several factors, including the type of intervention, the delay before implementation, and the intervention’s effectiveness \citep{longiniContainingPandemicInfluenza2005}. For highly transmissible pathogens with short generation times (e.g. influenza), swift and effective interventions are required to achieve containment. Pathogens that have a large proportion of onward transmission prior to symptom onset (latent period shorter than incubation period) or have asymptomatic (subclinical) cases are less easily controlled \citep{fraserFactorsThatMake2004}. \\

Isolation of symptomatic infections aims to break the chain of transmission by preventing further infection of susceptible individuals. The earlier an infectious case can be isolated, the more effective it is at averting secondary cases, with isolation prior to becoming infectious optimal. This can make it challenging to control outbreaks with symptom-based isolation, as for most pathogens infectiousness preceeds prodromal symptoms. Quarantining contacts exposed to an infectious individuals, independent of whether they later become a case, is a control strategy to reduce transmission before or shortly after symptom onset. Contact tracing is the process of finding the contacts of infected individuals. Both isolation and contact tracing are often used NPIs to try and curtail the growth of an epidemic. Forward contact tracing asks infected individuals who they have been in contact with so their contacts can be followed-up and isolated or quarantined if needed. Isolation, quarantine, and contact tracing is a targeted approach that looks to prevent the onward transmission of an infectious disease by tackling infected individuals and their contacts \citep{ferrettiQuantifyingSARSCoV2Transmission2020, keelingEfficacyContactTracing2020}. Isolation or vaccination of traced individuals has proved widely effective in controlling different outbreaks with a variety of pathogens differing in their aetiological and epidemiological characteristics \citep{foegeSELECTIVEEPIDEMIOLOGICCONTROL1971, bellPublicHealthInterventions2004}. However, in epidemics or pandemics where the proportion of cases or contacts missed is high, either because surveillance is patchy or because the number of contacts exceeds the capacity of the system, the efficacy of contact tracing to contain an outbreak diminishes, or at worst, fails completely \citep{dhillonWhenContactTracing2018, hellewellFeasibilityControllingCOVID192020}. \\

Several pre-COVID-19 studies have concluded that influenza viruses cannot be controlled by isolating individuals at symptom onset and contact tracing, because the rate at which infected individuals become infectious and transmit is faster than the latency in the process of isolating contacts \citep{fraserFactorsThatMake2004, klinkenbergEffectivenessContactTracing2006}. However, with the recent human infections of avian influenza (A/H5N1), mostly among poultry and cattle farmers \citep{gargHighlyPathogenicAvian2025} and the longer estimates of incubation period and serial interval from historical H5N1 data \citep{Ward2024.12.11.24318702} it may be that isolation and contact tracing is able to control influenza subtypes with pandemic potential. The (antigenic) novelty of a pathogen introduced into a human population is a leading risk factor for causing a pandemic. Thus, influenza subtypes that have low levels of existing immunity in the population are probable epidemic and pandemic candidates. This is the case for H5N1 \citep{Ward2024.12.11.24318702} and H7N9 \citep{tannerPandemicPotentialAvian2015} and was the case when H1N1 caused an outbreak in 2009 \citep{fraserPandemicPotentialStrain2009}. Historically, avian influenza (H5N1 and H7N9) has had high severity (case fatality risk $\sim 35\%-55\%$), and therefore sustained transmission in a human population could have substantial mortality and morbidity implications \citep{tannerPandemicPotentialAvian2015}. \\

@@

\begin{figure}[ht]
\centering
\includegraphics[width=\textwidth]{../plots/model-schematic.png}
\caption{Model schematic of the branching process with isolation, quarantine and contact tracing implemented in the \texttt{\{ringbp\}} R package. (A) The transmission chain from the index case (Person A), infection events are shown by black vertical dashed lines. The time individuals are infected is shown by the red dot, the time of symptom onset is shown by the orange dot, and the time time individuals are isolated is shown by the blue dot. Quarantine is optional in the model, assuming quarantine is active it is shown by the small green dot with a Q. The time at whcih a infectors isolation triggers the isolation of a symtomatic case is shown by the blue vertical dashed line (Person A to Person B). The isolation times that would have been triggered after a delay from symptom onset, but are after isolation was triggered by contact tracing shown by a blue dot with a cross and connected by a dotted line (Person B). Person C is an asymptomatic individual (asymptomatic cases can infect others). Person D is missed by contact tracing (contacts are traced with a probability $\rho$, and thus missed at $1 - \rho$). The time Person D would have isolated if ascertained is shown by the vertical blue mark. Assuming quarantine is active, Person F would be isolated before symptom onset at the time of the infector's (Person D) isolation time. If quarantine is not active, then Person F would be isolated after a delay from symptom onset, as they had yet to show symptoms when the infector was isolated. (B) A branching process diagram showing who-infected-whom.}
\label{fig:ringbp-model}
\end{figure}

The time between an infector being infected and then infecting an infectee (i.e. generation time) is computed using the incubation period (which samples the symptom onset time from the exposure time of each case) and sampling from a skew-normal distribution \citep{azzaliniClassDistributionsWhich1985}. This approach enables simulating transmission dynamics with a correlated structured between each case's incubation period and generation time for sensible delays between exposure, onset of symptoms, and onward transmission \citep{hellewellFeasibilityControllingCOVID192020}. This allows the proportion presymptomatic transmission to be specified, rather than defining incubation period and generation time distributions, which only indirectly allows assumptions of presymptomatic infectiousness to be derived. This simple formulation is advantageous for novel pathogens and early in outbreaks when the incubation period is estimated but the generation time is estimated with more uncertainty or unknown, and the serial interval cannot be substituted in without introducing bias \citep{brittonEstimationEmergingEpidemics2019a, lehtinenRelationshipSerialInterval2021}. \\

Given the incubation period for each case, the proportion of the infections that occur prior to, or after, the symptom onset of an infector is controlled by the parameterisation of the skew normal distribution (Figure \ref{fig:delay-distributions}C). The symptom onset time is used as the location parameter ($\xi$) of the distribution, providing the point around which the distribution is dispersed. The scale parameter ($\omega$) is fixed at 2, this parameter (with the shape parameter) defines the variance of the distribution. We select this value as it provides a realistic amount of variance when transforming onset time to exposure time (variance range for scenarios in this study: 1.5 --- 3.2). The shape parameter ($\alpha$) defines the skewness of the distribution, whereby positive values cause right-skewed values distribution around the location $\xi$, and negative values have the opposite effect (as $\alpha \rightarrow \infty$, the skew-normal converges to a folded-normal distribution). In summary, for a given parameterisation of the skew-normal distribution given the time of symptom onset ($\xi$) and fixing $\omega$ and $\alpha$ we sample the time an infectee is exposed/infected by an infector. \\

@@
